---
title: "Intial Methylation Assessment"
author: "Yaamini Venkataraman"
date: "03/04/2021"
output: github_document
---

In this script, I'll get preliminary methylation information from diploid and triploid oyster samples. Additionally, identify differentially methylated loci (DML) between treatments and ploidy status using pairwise comparisons.

# Prepare R Markdown file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install packages

```{r}
install.packages("devtools") #Install the devtools package
require(devtools) #Load devtools
```

```{r}
source("http://bioconductor.org/biocLite.R") #Source package from bioconductor
BiocManager::install("methylKit") #Install methylkit
```

```{r}
install_github("al2na/methylKit", build_vignettes = FALSE, repos = BiocInstaller::biocinstallRepos(), dependencies = TRUE) #Install more methylKit options
```

```{r}
require(methylKit) #Load methylkit
```

```{r}
require(dplyr) #Load dplyr
```

# Obtain session information

```{r}
sessionInfo()
```

# Download files from `gannet`

```{bash}
wget -r -l1 --no-parent --no-directories -A.merged_CpG_evidence.cov https://gannet.fish.washington.edu/spartina/project-oyster-oa/Haws/bismark/ #Download files from gannet to the curent directory (where script is saved). Exclude directory structure
```

```{bash}
mv *.merged_CpG_evidence.cov ../../analyses/Haws_04-methylKit #Move files from current directory to analyses folder
```

# Process methylation data

```{r}
analysisFiles <- list("zr3644_1_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_2_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_3_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_4_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_5_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_6_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_7_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_8_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_9_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_10_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_11_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_12_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_13_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_14_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_15_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_16_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_17_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_18_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_19_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_20_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_21_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_22_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_23_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "zr3644_24_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov") #Put all .cov files into a list for analysis.
```

I'll use `methRead` to create a methylation object from the coverage files, and include sample ID and treatment information.

```{r}
setwd(dir = "/Users/yaamini/Documents/project-oyster-oa/analyses/Haws_04-methylKit/") #Set working directory inside code chunk
processedFiles <- methylKit::methRead(analysisFiles,
                                      sample.id = list("2H-1", "2H-2", "2H-3", "2H-4", "2H-5", "2H-6",
                                                       "2L-1", "2L-2", "2L-3", "2L-4", "2L-5", "2L-6",
                                                       "3H-1", "3H-2", "3H-3", "3H-4", "3H-5", "3H-6",
                                                       "3L-1", "3L-2", "3L-3", "3L-4", "3L-5", "3L-6"),
                                      assembly = "oyster_v9",
                                      treatment = c(rep(0, times = 12),
                                                    rep(1, times = 12)),
                                      pipeline = "bismarkCoverage",
                                      mincov = 2) #Process files. Treatment specified based on ploidy status. Use mincov = 2 to quickly process reads.
```

```{r}
#save.image("../../analyses/Haws_04-methylKit/methylKit.RData") #Save R Data in case R crashes
load("../../analyses/Haws_04-methylKit/methylKit.RData") #Load R Data
```

```{r}
processedFilteredFilesCov5 <- methylKit::filterByCoverage(processedFiles,
                                                          lo.count = 5, lo.perc = NULL,
                                                          high.count = NULL, high.perc = 99.9) %>% 
  methylKit::normalizeCoverage(processedFilteredFilesCov5) #Filter coverage information for minimum 5x coverage, and remove PCR duplicates by excluding data in the 99.9th percentile of coverage with hi.perc = 99.9. Normalize coverage between samples to avoid over-sampling reads from one sample during statistical testing
```

```{r}
save.image("../../analyses/Haws_04-methylKit/methylKit.RData") #Save R Data in case R crashes
#load("../../analyses/Haws_04-methylKit/methylKit.RData") #Load R Data
```

# Characterize general methylation

## Sample-specific descriptive statistics

### Specify working directory for output

```{bash}
mkdir ../../analyses/Haws_04-methylKit/general-stats
```

```{r}
nFiles <- 24 #Count number of samples
fileName <- data.frame("nameBase" = rep("../../analyses/Haws_04-methylKit/general-stats/percent-CpG-methylation", times = nFiles),
                       "nameBase2" = rep("../../analyses/Haws_04-methylKit/general-stats/percent-CpG-coverage", times = nFiles),
                       "sample.ID" = 1:24) #Create new dataframe for filenames
head(fileName) #Confirm dataframe creation
```

```{r}
fileName$actualFileName1 <- paste(fileName$nameBase, "-Filtered", "-5xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for filtered + 5x coverage + specific sample's percent CpG methylation plot
fileName$actualFileName2 <- paste(fileName$nameBase2, "-Filtered", "-5xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for filtered + 5x coverage + specific sample's percent CpG coverage plot
head(fileName) #Confirm column creation
```

### Create plots

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName1[i], height = 1000, width = 1000) #Save file with designated name
  methylKit::getMethylationStats(processedFilteredFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get %CpG methylation information
  dev.off() #Turn off plotting device
} #Plot and save %CpG methylation information
```

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName2[i], height = 1000, width = 1000) #Save file with designated name
  methylKit::getCoverageStats(processedFilteredFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get CpG coverage information
  dev.off() #Turn off plotting device
} #Plot and save CpG coverage information
```

## Comparative analysis

```{r}
methylationInformationFilteredCov5 <- methylKit::unite(processedFilteredFilesCov5, 
                                                       destrand = FALSE) #Combine all processed files into a single table. Use destrand = TRUE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5) #Confirm unite
```

```{r}
save.image("../../analyses/Haws_04-methylKit/methylKit.RData") #Save R Data in case R crashes
#load("../../analyses/Haws_04-methylKit/methylKit.RData") #Load R Data
```

```{r}
clusteringInformationFilteredCov5 <- methylKit::clusterSamples(methylationInformationFilteredCov5, dist = "correlation", method = "ward", plot = FALSE) #Save cluster information as a new object
```

```{r}
jpeg(filename = "../../analyses/Haws_04-methylKit/general-stats/Full-Sample-Pearson-Correlation-Plot-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::getCorrelation(methylationInformationFilteredCov5, plot = TRUE) #Understand correlation between methylation patterns in different samples
dev.off()
```

```{r}
jpeg(filename = "../../analyses/Haws_04-methylKit/general-stats/Full-Sample-CpG-Methylation-Clustering-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::clusterSamples(methylationInformationFilteredCov5, dist = "correlation", method = "ward", plot = TRUE) #Cluster samples based on correlation coefficients
dev.off()
```

```{r}
jpeg(filename = "../../analyses/Haws_04-methylKit/general-stats/Full-Sample-Methylation-PCA-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::PCASamples(methylationInformationFilteredCov5) #Run a PCA analysis on percent methylation for all samples
dev.off() #Turn off plotting device
```

```{r}
jpeg(filename = "../../analyses/Haws_04-methylKit/general-stats/Full-Sample-Methylation-Screeplot-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::PCASamples(methylationInformationFilteredCov5, screeplot = TRUE) #Run the PCA analysis and plot variances against PC number in a screeplot
dev.off()
```

```{r}
save.image("../../analyses/Haws_04-methylKit/methylKit.RData") #Save R Data in case R crashes
#load("../../analyses/Haws_04-methylKit/methylKit.RData") #Load R Data
```

# Differentially methylated loci

I will examine ploidy and treatment differences separately, using the other variable as a covariate.

## Create covariate matrix

Each covaraite needs its own matrix, based on [the user guide](https://bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html#38_Accounting_for_covariates).

```{r}
covariatepH <- data.frame("pH" = c(rep("H", times = 6),
                                 rep("L", times = 6),
                                 rep("H", times = 6),
                                 rep("L", times = 6))) #Create dataframe with pH covariate information
head(covariatepH) #Check dataframe format
```

```{r}
covariatePloidy <- data.frame("ploidy" = c(rep("2N", times = 12), 
                                           rep("3N", times = 12))) #Create dataframe with ploidy covariate information
head(covariatePloidy) #Check dataframe format
```
   
## Ploidy differences

```{r}
diffMethStatsPloidy50 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5,
                                                                   covariates = covariatepH,
                                                                   mc.cores = 4) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate. Use 4 cores.
head(differentialMethylationStatsPloidy) #Look at differential methylation statistics
```

```{r}
diffMethStatsPloidy50 <- methylKit::getMethylDiff(differentialMethylationStatsPloidy, difference = 50, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatsPloidy50$chr) #Count the number of DML
head(diffMethStatsPloidy50) #Confirm creation
```

```{r}
diffMethStatsPloidy75 <- methylKit::getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 75, qvalue = 0.01) #Identify loci that are at least 75% different
length(diffMethStatsPloidy75$chr) #Count the number of DML
head(diffMethStatsPloidy75) #Confirm creation
```

```{r}
diffMethStatsPloidy100 <- methylKit::getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 99, qvalue = 0.01) #Identify loci that are at least 100% different
length(diffMethStatsPloidy100$chr) #Count the number of DML
head(diffMethStatsPloidy100) #Confirm creation
```

```{r}
save.image("../../analyses/Haws_04-methylKit/methylKit.RData") #Save R Data in case R crashes
#load("../../analyses/Haws_04-methylKit/methylKit.RData") #Load R Data
```

I'm going to use the following settings moving forward: 500% difference and q-value < 0.001.

```{r}
#write.csv(diffMethStatsPloidy50, "DML-ploidy-50-Cov5.csv") #Save table as .csv
```

#### Hypermethylated DML

```{r}
diffMethStatsPloidy50Hyper <- getMethylDiff(diffMethStatsPloidy50, difference = 50, qvalue = 0.01, type = "hyper") #Identify hypermethylated loci that are at least 50% different
head(diffMethStatsPloidy50Hyper) #Confirm creation
```

```{r}
#write.csv(diffMethStatsPloidy50Hyper, "DML-ploidy-50-Cov5-Hyper.csv") #Save table as .csv
```

#### Hypomethylated DML

```{r}
diffMethStatsPloidy50Hypo <- getMethylDiff(diffMethStatsPloidy50, difference = 50, qvalue = 0.01, type = "hyper") #Identify hypomethylated loci that are at least 50% different
head(diffMethStatsPloidy50Hypo) #Confirm creation
```

```{r}
#write.csv(diffMethStatsPloidy50Hypo, "DML-ploidy-50-Cov5-Hypo.csv") #Save table as .csv
```

## Treatment differences

### Assign treatment information

```{r}
processedFilesFilteredT <- methylKit::reorganize(processedFiles,
                                        sample.id = list("2H-1", "2H-2", "2H-3", "2H-4", "2H-5", "2H-6",
                                                         "2L-1", "2L-2", "2L-3", "2L-4", "2L-5", "2L-6",
                                                         "3H-1", "3H-2", "3H-3", "3H-4", "3H-5", "3H-6",
                                                         "3L-1", "3L-2", "3L-3", "3L-4", "3L-5", "3L-6"),
                                        treatment = c(rep(0, times = 6),
                                                      rep(1, times = 6),
                                                      rep(0, times = 6),
                                                      rep(1, times = 6))) %>%
  methylKit::filterByCoverage(.,
                              lo.count = 5, lo.perc = NULL,
                              high.count = NULL, high.perc = 99.9) %>% 
  methylKit::no
#Reorganize processedFiles to provide treatment specification. Filter by coverage and normalize coverage between samples
head(processedFilesT)
```








# Save DMLs as BEDfiles

## Install packages

**Do not** install these packages until analysis in `methylKit` is complete, as some of the installed packages will mask others important for analysis.

```{r}
require(readr) #Load package

install.packages("tidyverse") #Install tidyverse
require(tidyverse) #Load package
```

## DML

```{r}
DML5xFilteredDestrand09152019 <- mutate(diffMethStats50FilteredCov5Destrand, start = start - 3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML5xFilteredDestrand09152019) #Confirm changes
```

```{r}
write_delim(DML5xFilteredDestrand09152019, "2019-09-15-DML-Destrand-5x-Locations.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypermethylated DML

```{r}
DML5xFilteredDestrandHyper09152019 <- mutate(diffMethStats50FilteredCov5DestrandHyper, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML5xFilteredDestrandHyper09152019) #Confirm changes
```

```{r}
write_delim(DML5xFilteredDestrandHyper09152019, "2019-09-15-DML-Destrand-5x-Locations-Hypermethylated.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypomethyalted DML

```{r}
DML5xFilteredDestrandHypo09152019 <- mutate(diffMethStats50FilteredCov5DestrandHypo, start = start - 3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML5xFilteredDestrandHypo09152019) #Confirm changes
```

```{r}
write_delim(DML5xFilteredDestrandHypo09152019, "2019-09-15-DML-Destrand-5x-Locations-Hypomethylated.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```
