---
title: "Methylation Parameter Testing"
author: "Yaamini Venkataraman"
date: "07/10/2026"
output: github_document
---

In this script, I will assess different parameters for DML identification.

# Prepare R Markdown file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dir.create("../../analyses/Haws_04.2-methylKit/DML", recursive = TRUE, showWarnings = FALSE) #Create output directories if they do not exist. normalizePath() below silently returns an unresolved relative path if the target is missing, which breaks every relative output path in this script
dir.create("../../analyses/Haws_04.2-methylKit/general-stats", recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../../analyses/Haws_04.2-methylKit/")) #Set root directory
```

```{r}
getwd()
```




# Install packages

```{r, include = FALSE}
# install.packages("tidyverse")
# install.packages("data.table") #Personal computer
# install.packages("https://cran.r-project.org/src/contrib/Archive/data.table/data.table_1.16.4.tar.gz",
#                  repos = NULL,
#                  type = "source",
#                  lib = "/home/shared/8TB_HDD_02/yaaminiv/R/x86_64-pc-linux-gnu-library/4.2") # R Studio server: need legacy data.table in order to install legacy methylKit
# install.packages("devtools")
# install.packages("BiocManager")
# install.packages("vegan")
# install.packages("scales")
# install.packages("RColorBrewer")
# install.packages("pheatmap")
```

```{r, include = FALSE}
require(tidyverse)
require(data.table)
require(BiocManager)
# BiocManager::install("methylKit") #Install methylKit
require(methylKit) #Load methylKit
require(vegan)
require(scales)
require(RColorBrewer)
require(pheatmap)
```

# Obtain session information

```{r}
sessionInfo()
```

# Process methylation data

```{r}
analysisFiles <- list("../../data/Haws/zr3644_1_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_2_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_3_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_4_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_5_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_6_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_7_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_8_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_9_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_10_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_11_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_12_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_13_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_14_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_15_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_16_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_17_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_18_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_19_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_20_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_21_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_22_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_23_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                      "../../data/Haws/zr3644_24_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov") #Put all .cov files into a list for analysis.
```

```{r}
sampleMetadata <- read.csv("../../data/Haws/sample_metadata.csv") %>%
  dplyr::select(-c(1)) %>%
  arrange(sample_number) %>%
  mutate("sampleID" = c("2H-1", "2H-2", "2H-3", "2H-4", "2H-5", "2H-6",
                        "2L-1", "2L-2", "2L-3", "2L-4", "2L-5", "2L-6",
                        "3H-1", "3H-2", "3H-3", "3H-4", "3H-5", "3H-6",
                        "3L-1", "3L-2", "3L-3", "3L-4", "3L-5", "3L-6")) %>%
  dplyr::rename(pH = ph) %>%
  mutate(ploidyTreatment = case_when(ploidy == "3N" ~ 1,
                                     ploidy == "2N" ~ 0)) %>%
  mutate(pHTreatment = case_when(pH == "high" ~ 0,
                                 pH == "low" ~ 1)) #Import metadata and modify for use with methylKit
head(sampleMetadata) #Confirm dataframe creation
```

I'll use `methRead` to create a methylation object from the coverage files, and include sample ID and treatment information.

```{r}
processedFiles <- methylKit::methRead(analysisFiles,
                                      sample.id = list("2H-1", "2H-2", "2H-3", "2H-4", "2H-5", "2H-6",
                                                       "2L-1", "2L-2", "2L-3", "2L-4", "2L-5", "2L-6",
                                                       "3H-1", "3H-2", "3H-3", "3H-4", "3H-5", "3H-6",
                                                       "3L-1", "3L-2", "3L-3", "3L-4", "3L-5", "3L-6"),
                                      assembly = "oyster_v9",
                                      treatment = sampleMetadata$ploidyTreatment,
                                      pipeline = "bismarkCoverage",
                                      mincov = 2) #Process files. Treatment specified based on ploidy status. Use mincov = 2 to quickly process reads.
```

```{r}
processedFilteredFilesCov5 <- methylKit::filterByCoverage(processedFiles,
                                                          lo.count = 5, lo.perc = NULL,
                                                          high.count = NULL, high.perc = 99.9) %>%
  methylKit::normalizeCoverage(.) #Filter coverage information for minimum 5x coverage, and remove PCR duplicates by excluding data in the 99.9th percentile of coverage with hi.perc = 99.9. Normalize coverage between samples to avoid over-sampling reads from one sample during statistical testing
```

```{r}
save.image("methylKit.RData") #Save R Data in case R crashes
#load("methylKit.RData") #Load R Data
```

# Testing `min.per.group`

Modifying `min.per.group` will determine how many samples a specific CpG locus needs at least 5x coverage in to be used for DML calling.

## Ploidy

```{r}
methylationInformationFilteredCov5PloidyAll <- methylKit::unite(processedFilteredFilesCov5,
                                                                destrand = FALSE,
                                                                mc.cores = 2) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5PloidyAll) #Confirm unite
```

```{r}
methylationInformationFilteredCov5Ploidy11 <- methylKit::unite(processedFilteredFilesCov5,
                                                               destrand = FALSE,
                                                               mc.cores = 2,
                                                               min.per.group = 11L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5Ploidy11) #Confirm unite
```

```{r}
methylationInformationFilteredCov5Ploidy10 <- methylKit::unite(processedFilteredFilesCov5,
                                                               destrand = FALSE,
                                                               mc.cores = 2,
                                                               min.per.group = 10L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5Ploidy10) #Confirm unite
```

```{r}
methylationInformationFilteredCov5Ploidy9 <- methylKit::unite(processedFilteredFilesCov5,
                                                              destrand = FALSE,
                                                              mc.cores = 2,
                                                              min.per.group = 9L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5Ploidy9) #Confirm unite
```

```{r}
methylationInformationFilteredCov5Ploidy8 <- methylKit::unite(processedFilteredFilesCov5,
                                                              destrand = FALSE,
                                                              mc.cores = 2,
                                                              min.per.group = 8L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5Ploidy8) #Confirm unite
```

## pH

```{r}
methylationInformationFilteredCov5pHAll <- methylKit::reorganize(methylationInformationFilteredCov5PloidyAll,
                                                                 sample.id = sampleMetadata$sampleID,
                                                                 treatment = sampleMetadata$pHTreatment) #Reorganize methylationInformationFilteredCov5PloidyAll to provide pH treatment specification (can use only when not modifying min.per.group)
head(methylationInformationFilteredCov5pHAll) #Confirm unite
```

```{r}
processedFilespH <- methylKit::methRead(analysisFiles,
                                        sample.id = list("2H-1", "2H-2", "2H-3", "2H-4", "2H-5", "2H-6",
                                                         "2L-1", "2L-2", "2L-3", "2L-4", "2L-5", "2L-6",
                                                         "3H-1", "3H-2", "3H-3", "3H-4", "3H-5", "3H-6",
                                                         "3L-1", "3L-2", "3L-3", "3L-4", "3L-5", "3L-6"),
                                        assembly = "oyster_v9",
                                        treatment = sampleMetadata$pHTreatment,
                                        pipeline = "bismarkCoverage",
                                        mincov = 2) #Process files. Treatment specified based on pH status. Use mincov = 2 to quickly process reads.
```

```{r}
processedFilteredFilesCov5pH <- methylKit::filterByCoverage(processedFilespH,
                                                            lo.count = 5, lo.perc = NULL,
                                                            high.count = NULL, high.perc = 99.9) %>%
  methylKit::normalizeCoverage(.) #Filter coverage information for minimum 5x coverage, and remove PCR duplicates by excluding data in the 99.9th percentile of coverage with hi.perc = 99.9. Normalize coverage between samples to avoid over-sampling reads from one sample during statistical testing
```

```{r}
methylationInformationFilteredCov5pH11 <- methylKit::unite(processedFilteredFilesCov5pH,
                                                           destrand = FALSE,
                                                           mc.cores = 2,
                                                           min.per.group = 11L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5pH11) #Confirm unite
```

```{r}
methylationInformationFilteredCov5pH10 <- methylKit::unite(processedFilteredFilesCov5pH,
                                                           destrand = FALSE,
                                                           mc.cores = 2,
                                                           min.per.group = 10L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5pH10) #Confirm unite
```

```{r}
methylationInformationFilteredCov5pH9 <- methylKit::unite(processedFilteredFilesCov5pH,
                                                          destrand = FALSE,
                                                          mc.cores = 2,
                                                          min.per.group = 9L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5pH9) #Confirm unite
```

```{r}
methylationInformationFilteredCov5pH8 <- methylKit::unite(processedFilteredFilesCov5pH,
                                                          destrand = FALSE,
                                                          mc.cores = 2,
                                                          min.per.group = 8L) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredCov5pH8) #Confirm unite
```

# Identify differentially methylated loci

## Ploidy differences

### All samples

```{r}
differentialMethylationStatsPloidyCovarAll <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5PloidyAll,
                                                                           covariates = as.data.frame(sampleMetadata$pH),
                                                                           mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(differentialMethylationStatsPloidyCovarAll) #Look at differential methylation statistics
```

```{r}
diffMethStatsPloidyCovar25All <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovarAll, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatsPloidyCovar25All$chr) #Count the number of DML
head(diffMethStatsPloidyCovar25All) #Confirm creation
```

```{r}
diffMethStatsPloidyCovar50All <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovarAll, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatsPloidyCovar50All$chr) #Count the number of DML
head(diffMethStatsPloidyCovar50All) #Confirm creation
```

```{r}

write_delim(diffMethStatsPloidyCovar25All, "DML/DML-ploidy-25-Covar-Cov5-All.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatsPloidyCovar50All, "DML/DML-ploidy-50-Covar-Cov5-All.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 11

```{r}
differentialMethylationStatsPloidyCovar11 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5Ploidy11,
                                                                          covariates = as.data.frame(sampleMetadata$pH),
                                                                          mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(differentialMethylationStatsPloidyCovar11) #Look at differential methylation statistics
```

```{r}
diffMethStatsPloidyCovar2511 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar11, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatsPloidyCovar2511$chr) #Count the number of DML
head(diffMethStatsPloidyCovar2511) #Confirm creation
```

```{r}
diffMethStatsPloidyCovar5011 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar11, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatsPloidyCovar5011$chr) #Count the number of DML
head(diffMethStatsPloidyCovar5011) #Confirm creation
```

```{r}
write_delim(diffMethStatsPloidyCovar2511, "DML/DML-ploidy-25-Covar-Cov5-11.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatsPloidyCovar5011, "DML/DML-ploidy-50-Covar-Cov5-11.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 10

```{r}
differentialMethylationStatsPloidyCovar10 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5Ploidy10,
                                                                          covariates = as.data.frame(sampleMetadata$pH),
                                                                          mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(differentialMethylationStatsPloidyCovar10) #Look at differential methylation statistics
```

```{r}
diffMethStatsPloidyCovar2510 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar10, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatsPloidyCovar2510$chr) #Count the number of DML
head(diffMethStatsPloidyCovar2510) #Confirm creation
```

```{r}
diffMethStatsPloidyCovar5010 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar10, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatsPloidyCovar5010$chr) #Count the number of DML
head(diffMethStatsPloidyCovar5010) #Confirm creation
```

```{r}
write_delim(diffMethStatsPloidyCovar2510, "DML/DML-ploidy-25-Covar-Cov5-10.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatsPloidyCovar5010, "DML/DML-ploidy-50-Covar-Cov5-10.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 9

```{r}
differentialMethylationStatsPloidyCovar9 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5Ploidy9,
                                                                         covariates = as.data.frame(sampleMetadata$pH),
                                                                         mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(differentialMethylationStatsPloidyCovar9) #Look at differential methylation statistics
```

```{r}
diffMethStatsPloidyCovar259 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar9, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatsPloidyCovar259$chr) #Count the number of DML
head(diffMethStatsPloidyCovar259) #Confirm creation
```

```{r}
diffMethStatsPloidyCovar509 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar9, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatsPloidyCovar509$chr) #Count the number of DML
head(diffMethStatsPloidyCovar509) #Confirm creation
```

```{r}
write_delim(diffMethStatsPloidyCovar259, "DML/DML-ploidy-25-Covar-Cov5-9.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatsPloidyCovar509, "DML/DML-ploidy-50-Covar-Cov5-9.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 8

```{r}
differentialMethylationStatsPloidyCovar8 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5Ploidy8,
                                                                         covariates = as.data.frame(sampleMetadata$pH),
                                                                         mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(differentialMethylationStatsPloidyCovar8) #Look at differential methylation statistics
```

```{r}
diffMethStatsPloidyCovar258 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar8, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatsPloidyCovar258$chr) #Count the number of DML
head(diffMethStatsPloidyCovar258) #Confirm creation
```

```{r}
diffMethStatsPloidyCovar508 <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovar8, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatsPloidyCovar508$chr) #Count the number of DML
head(diffMethStatsPloidyCovar508) #Confirm creation
```

```{r}
write_delim(diffMethStatsPloidyCovar258, "DML/DML-ploidy-25-Covar-Cov5-8.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatsPloidyCovar508, "DML/DML-ploidy-50-Covar-Cov5-8.bed", delim = "\t", quote = "none") #Save table a BED file
```

```{r}
save.image("methylKit.RData") #Save R Data in case R crashes
#load("methylKit.RData") #Load R Data
```

## pH differences

### All samples

```{r}
methylationInformationFilteredCov5pHAll <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5pHAll,
                                                                        covariates = as.data.frame(sampleMetadata$ploidy),
                                                                        mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(methylationInformationFilteredCov5pHAll) #Look at differential methylation statistics
```

```{r}
diffMethStatspHCovar25All <- methylKit::getMethylDiff(methylationInformationFilteredCov5pHAll, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatspHCovar25All$chr) #Count the number of DML
head(diffMethStatspHCovar25All) #Confirm creation
```

```{r}
diffMethStatspHCovar50All <- methylKit::getMethylDiff(methylationInformationFilteredCov5pHAll, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatspHCovar50All$chr) #Count the number of DML
head(diffMethStatspHCovar50All) #Confirm creation
```

```{r}
write_delim(diffMethStatspHCovar25All, "DML/DML-pH-25-Covar-Cov5-All.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatspHCovar50All, "DML/DML-pH-50-Covar-Cov5-All.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 11

```{r}
methylationInformationFilteredCov5pH11 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5pH11,
                                                                       covariates = as.data.frame(sampleMetadata$ploidy),
                                                                       mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(methylationInformationFilteredCov5pH11) #Look at differential methylation statistics
```

```{r}
diffMethStatspHCovar2511 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH11, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatspHCovar2511$chr) #Count the number of DML
head(diffMethStatspHCovar2511) #Confirm creation
```

```{r}
diffMethStatspHCovar5011 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH11, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatspHCovar5011$chr) #Count the number of DML
head(diffMethStatspHCovar5011) #Confirm creation
```

```{r}
write_delim(diffMethStatspHCovar2511, "DML/DML-pH-25-Covar-Cov5-11.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatspHCovar5011, "DML/DML-pH-50-Covar-Cov5-11.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 10

```{r}
methylationInformationFilteredCov5pH10 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5pH10,
                                                                       covariates = as.data.frame(sampleMetadata$ploidy),
                                                                       mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(methylationInformationFilteredCov5pH10) #Look at differential methylation statistics
```

```{r}
diffMethStatspHCovar2510 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH10, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatspHCovar2510$chr) #Count the number of DML
head(diffMethStatspHCovar2510) #Confirm creation
```

```{r}
diffMethStatspHCovar5010 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH10, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatspHCovar5010$chr) #Count the number of DML
head(diffMethStatspHCovar5010) #Confirm creation
```

```{r}
write_delim(diffMethStatspHCovar2510, "DML/DML-pH-25-Covar-Cov5-10.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatspHCovar5010, "DML/DML-pH-50-Covar-Cov5-10.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 9

```{r}
methylationInformationFilteredCov5pH9 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5pH9,
                                                                      covariates = as.data.frame(sampleMetadata$ploidy),
                                                                      mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(methylationInformationFilteredCov5pH9) #Look at differential methylation statistics
```

```{r}
diffMethStatspHCovar259 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH9, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatspHCovar259$chr) #Count the number of DML
head(diffMethStatspHCovar259) #Confirm creation
```

```{r}
diffMethStatspHCovar509 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH9, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatspHCovar509$chr) #Count the number of DML
head(diffMethStatspHCovar509) #Confirm creation
```

```{r}
write_delim(diffMethStatspHCovar259, "DML/DML-pH-25-Covar-Cov5-9.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatspHCovar509, "DML/DML-pH-50-Covar-Cov5-9.bed", delim = "\t", quote = "none") #Save table a BED file
```

### `min.per.group` = 8

```{r}
methylationInformationFilteredCov5pH8 <- methylKit::calculateDiffMeth(methylationInformationFilteredCov5pH8,
                                                                      covariates = as.data.frame(sampleMetadata$ploidy),
                                                                      mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(methylationInformationFilteredCov5pH8) #Look at differential methylation statistics
```

```{r}
diffMethStatspHCovar258 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH8, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatspHCovar258$chr) #Count the number of DML
head(diffMethStatspHCovar258) #Confirm creation
```

```{r}
diffMethStatspHCovar508 <- methylKit::getMethylDiff(methylationInformationFilteredCov5pH8, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatspHCovar508$chr) #Count the number of DML
head(diffMethStatspHCovar508) #Confirm creation
```

```{r}
write_delim(diffMethStatspHCovar258, "DML/DML-pH-25-Covar-Cov5-8.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatspHCovar508, "DML/DML-pH-50-Covar-Cov5-8.bed", delim = "\t", quote = "none") #Save table a BED file
```

# Outlier removal testing

Samples 3 and 14 are likely outliers. I will determine how outlier removal impacts DML identification. I will do this using all samples with no `min.per.group` modification.

## Remove outliers from data

```{r}
analysisFilesOutRM <- list("../../data/Haws/zr3644_1_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_2_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_4_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_5_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_6_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_7_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_8_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_9_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_10_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_11_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_12_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_13_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_15_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_16_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_17_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_18_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_19_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_20_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_21_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_22_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_23_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov",
                           "../../data/Haws/zr3644_24_R1_val_1_val_1_val_1_bismark_bt2_pe..CpG_report.merged_CpG_evidence.cov") #Put all .cov files into a list for analysis.
```

```{r}
processedFilesOutRM <- methylKit::methRead(analysisFilesOutRM,
                                           sample.id = list("2H-1", "2H-2", "2H-4", "2H-5", "2H-6",
                                                            "2L-1", "2L-2", "2L-3", "2L-4", "2L-5", "2L-6",
                                                            "3H-1", "3H-3", "3H-4", "3H-5", "3H-6",
                                                            "3L-1", "3L-2", "3L-3", "3L-4", "3L-5", "3L-6"),
                                           assembly = "roslin",
                                           treatment = sampleMetadata$ploidyTreatment[-c(3,14)],
                                           pipeline = "bismarkCoverage",
                                           mincov = 2) #Process files. Treatment specified based on ploidy status. Use mincov = 2 to quickly process reads.
```

```{r}
processedFilesOutRMCov5 <- methylKit::filterByCoverage(processedFilesOutRM,
                                                       lo.count = 5, lo.perc = NULL,
                                                       high.count = NULL, high.perc = 99.9) %>%
  methylKit::normalizeCoverage(.) #Filter coverage information for minimum 5x coverage, and remove PCR duplicates by excluding data in the 99.9th percentile of coverage with hi.perc = 99.9. Normalize coverage between samples to avoid over-sampling reads from one sample during statistical testing
```

```{r}
save.image("methylKit.RData") #Save R Data in case R crashes
#load("methylKit.RData") #Load R Data
```

## Remake clustering and PCA

```{r}
methylationInformationFilteredRMCov5 <- methylKit::unite(processedFilesOutRMCov5,
                                                         destrand = FALSE,
                                                         mc.cores = 2) #Combine all processed files into a single table. Use destrand = FALSE to not destrand. By default only bases with data in all samples will be kept
head(methylationInformationFilteredRMCov5) #Confirm unite
```

```{r}
jpeg(filename = "general-stats/Full-Sample-Pearson-Correlation-Plot-FilteredCov5Destrand-OutRM.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::getCorrelation(methylationInformationFilteredRMCov5, plot = TRUE) #Understand correlation between methylation patterns in different samples
dev.off()
```

```{r}
jpeg(filename = "general-stats/Full-Sample-CpG-Methylation-Clustering-FilteredCov5Destrand-OutRM.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::clusterSamples(methylationInformationFilteredRMCov5, dist = "correlation", method = "ward", plot = TRUE) #Cluster samples based on correlation coefficients
dev.off()
```

```{r}
jpeg(filename = "general-stats/Full-Sample-Methylation-PCA-FilteredCov5Destrand-OutRM.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::PCASamples(methylationInformationFilteredRMCov5) #Run a PCA analysis on percent methylation for all samples
dev.off() #Turn off plotting device
```

```{r}
jpeg(filename = "general-stats/Full-Sample-Methylation-Screeplot-FilteredCov5Destrand-OutRM.jpeg", height = 1000, width = 1000) #Save file with designated name
methylKit::PCASamples(methylationInformationFilteredRMCov5, screeplot = TRUE) #Run the PCA analysis and plot variances against PC number in a screeplot
dev.off()
```

```{r}
save.image("methylKit.RData") #Save R Data in case R crashes
#load("methylKit.RData") #Load R Data
```

## Identify differentially methylated loci

### Ploidy DML

```{r}
differentialMethylationStatsPloidyCovarOutRMAll <- methylKit::calculateDiffMeth(methylationInformationFilteredRMCov5,
                                                                                covariates = as.data.frame(sampleMetadata$pH[-c(3,14)]),
                                                                                mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from methRead. Include pH as a covariate by converting column with pH information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(differentialMethylationStatsPloidyCovarOutRMAll) #Look at differential methylation statistics
```

```{r}
diffMethStatsPloidyCovar25RMAll <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovarOutRMAll, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatsPloidyCovar25RMAll$chr) #Count the number of DML
head(diffMethStatsPloidyCovar25RMAll) #Confirm creation
```

```{r}
diffMethStatsPloidyCovar50RMAll <- methylKit::getMethylDiff(differentialMethylationStatsPloidyCovarOutRMAll, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatsPloidyCovar50RMAll$chr) #Count the number of DML
head(diffMethStatsPloidyCovar50RMAll) #Confirm creation
```

```{r}
write_delim(diffMethStatsPloidyCovar25RMAll, "DML/DML-ploidy-25-Covar-Cov5-All-OutRM.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatsPloidyCovar50RMAll, "DML/DML-ploidy-50-Covar-Cov5-All-OutRM.bed", delim = "\t", quote = "none") #Save table a BED file
```

### pH DML

```{r}
methylationInformationFilteredRMCov5pH <- methylKit::reorganize(methylationInformationFilteredRMCov5,
                                                                sample.id = sampleMetadata$sampleID[-c(3,14)],
                                                                treatment = sampleMetadata$pHTreatment[-c(3,14)]) #Reorganize methylationInformationFilteredRMCov5 to provide pH treatment specification. methylationInformationFilteredRMCov5 was created with ploidy as the treatment, so it cannot be used directly for pH contrasts
head(methylationInformationFilteredRMCov5pH) #Confirm reorganize
```

```{r}
differentialMethylationStatspHCovarOutRMAll <- methylKit::calculateDiffMeth(methylationInformationFilteredRMCov5pH,
                                                                                covariates = as.data.frame(sampleMetadata$ploidy[-c(3,14)]),
                                                                                mc.cores = 40) #Calculate differential methylation statistics based on treatment indication from reorganize. Include ploidy as a covariate by converting column with ploidy information into a dataframe. Use an overdispersion correction with the F-test (default). Use 40 cores.
head(differentialMethylationStatspHCovarOutRMAll) #Look at differential methylation statistics
```

```{r}
diffMethStatspHCovar25RMAll <- methylKit::getMethylDiff(differentialMethylationStatspHCovarOutRMAll, difference = 25, qvalue = 0.01) #Identify loci that are at least 50% different
length(diffMethStatspHCovar25RMAll$chr) #Count the number of DML
head(diffMethStatspHCovar25RMAll) #Confirm creation
```

```{r}
diffMethStatspHCovar50RMAll <- methylKit::getMethylDiff(differentialMethylationStatspHCovarOutRMAll, difference = 50, qvalue = 0.01) #Identify loci that are at least 25% different
length(diffMethStatspHCovar50RMAll$chr) #Count the number of DML
head(diffMethStatspHCovar50RMAll) #Confirm creation
```

```{r}
write_delim(diffMethStatspHCovar25RMAll, "DML/DML-pH-25-Covar-Cov5-All-OutRM.bed", delim = "\t", quote = "none") #Save table a BED file
write_delim(diffMethStatspHCovar50RMAll, "DML/DML-pH-50-Covar-Cov5-All-OutRM.bed", delim = "\t", quote = "none") #Save table a BED file
```

# Summary table

```{r}
summaryDML <- data.frame("min.per.group" = c(rev(8:12), 12),
                         "ploidy-DML-25" = c(length(diffMethStatsPloidyCovar25All$chr),
                                             length(diffMethStatsPloidyCovar2511$chr),
                                             length(diffMethStatsPloidyCovar2510$chr),
                                             length(diffMethStatsPloidyCovar259$chr),
                                             length(diffMethStatsPloidyCovar258$chr),
                                             length(diffMethStatsPloidyCovar25RMAll$chr)),
                         "ploidy-DML-50" = c(length(diffMethStatsPloidyCovar50All$chr),
                                             length(diffMethStatsPloidyCovar5011$chr),
                                             length(diffMethStatsPloidyCovar5010$chr),
                                             length(diffMethStatsPloidyCovar509$chr),
                                             length(diffMethStatsPloidyCovar508$chr),
                                             length(diffMethStatsPloidyCovar50RMAll$chr)),
                         "pH-DML-25" = c(length(diffMethStatspHCovar25All$chr),
                                         length(diffMethStatspHCovar2511$chr),
                                         length(diffMethStatspHCovar2510$chr),
                                         length(diffMethStatspHCovar259$chr),
                                         length(diffMethStatspHCovar258$chr),
                                         length(diffMethStatspHCovar25RMAll$chr)),
                         "pH-DML-50" = c(length(diffMethStatspHCovar50All$chr),
                                         length(diffMethStatspHCovar5011$chr),
                                         length(diffMethStatspHCovar5010$chr),
                                         length(diffMethStatspHCovar509$chr),
                                         length(diffMethStatspHCovar508$chr),
                                         length(diffMethStatspHCovar50RMAll$chr))) #Create table with number of DML from each
```


