---
title: "General Methylation Landscape"
author: "Yaamini Venkataraman"
output: html_document
---

In this script, I'll create summary tables for genomic location information, run statistcal tests, and visualize the distribution of DML in the genome.

# Prepare R Markdown file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../analyses/Haws_06-methylation-landscape/")) #Set root directory
```

```{r}
getwd()
```

# Install packages

```{r}
#install.packages("broom")
#install.packages("RColorBrewer")
#install.packages("tidyverse")

library(broom)
library(RColorBrewer)
library(tidyverse)
```

# Obtain session information

```{r}
sessionInfo()
```

# Establish color scheme

```{r}
display.brewer.all()
plotColors <- brewer.pal(8, "Greys") #Create a color scheme
```

# Genomic location of methylated CpGs

My goal is to 1) understand statistical differences between distribution of CG motifs and methylated CpGs in various genomic features and 2) visualize the genomic locations of these loci.

## Import file counts

```{r}
exonUTROverlaps <- read.table("zr3644_5x-NO-SNPs-exonUTR-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
exonUTROverlaps <- exonUTROverlaps[73:75,] #Retain final rows
head(exonUTROverlaps) #Confirm import
```

```{r}
CDSOverlaps <- read.table("zr3644_5x-NO-SNPs-CDS-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
CDSOverlaps <- CDSOverlaps[73:75,] #Retain final rows
head(CDSOverlaps) #Confirm import
```

```{r}
intronOverlaps <- read.table("zr3644_5x-NO-SNPs-intron-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
intronOverlaps <- intronOverlaps[73:75,] #Retain final rows
head(intronOverlaps) #Confirm import
```

```{r}
upstreamOverlaps <- read.table("zr3644_5x-NO-SNPs-upstreamFlanks-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
upstreamOverlaps <- upstreamOverlaps[73:75,] #Retain final rows
head(upstreamOverlaps) #Confirm import
```

```{r}
downstreamOverlaps <- read.table("zr3644_5x-NO-SNPs-downstreamFlanks-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
downstreamOverlaps <- downstreamOverlaps[73:75,] #Retain final rows
head(downstreamOverlaps) #Confirm import
```

```{r}
lncRNAOverlaps <- read.table("zr3644_5x-NO-SNPs-lncRNA-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
lncRNAOverlaps <- lncRNAOverlaps[73:75,] #Retain final rows
head(lncRNAOverlaps) #Confirm import
```

```{r}
TEOverlaps <- read.table("zr3644_5x-NO-SNPs-TE-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
TEOverlaps <- TEOverlaps[73:75,] #Retain final rows
head(TEOverlaps) #Confirm import
```

```{r}
intergenicOverlaps <- read.table("zr3644_5x-NO-SNPs-intergenic-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
intergenicOverlaps <- intergenicOverlaps[73:75,] #Retain final rows
head(intergenicOverlaps) #Confirm import
```

## Create summary table

```{r}
cpgFeatureOverlaps <- cbind(exonUTROverlaps,
                            CDSOverlaps,
                            intronOverlaps,
                            upstreamOverlaps,
                            downstreamOverlaps,
                            lncRNAOverlaps,
                            TEOverlaps,
                            intergenicOverlaps) #Combine information from all genome features
rownames(cpgFeatureOverlaps) <- c("Meth", "sparseMeth", "unMeth") #Assign row names
ncol(cpgFeatureOverlaps) #Count columns
cpgFeatureOverlaps <- cpgFeatureOverlaps[,seq(1, 16, 2)] #Keep odd-numbered columns (counts)
colnames(cpgFeatureOverlaps) <- c("exonUTR", "CDS", "intron", "upstream", "downstream", "lncRNA", "TE", "intergenic") #Add column names
head(cpgFeatureOverlaps) #Confirm formatting
```

```{r}
cpgFeatureOverlaps <- as.data.frame(t(cpgFeatureOverlaps)) #Transpose so column names are row names
cpgFeatureOverlaps$allCpGs <- c(700167, 1516641, 5221932, 592834, 549787, 454393, 5424483, 4730728) #Add CpG motif overlap information
cpgFeatureOverlaps <- cpgFeatureOverlaps[,c(4, 1:3)] #Reorganize columns
head(cpgFeatureOverlaps) #Confirm formatting
```

```{r}
write.csv(cpgFeatureOverlaps, "CpG-feature-overlaps.csv", quote = FALSE, row.names = TRUE) #Save data
```

## Contingency tests

### Format data

```{r}
cpgLocationStatTest <- data.frame(t(cpgFeatureOverlaps)) #Transpose for statistical testing
cpgLocationStatTest <- cpgLocationStatTest[1:2,] #keep only allCpGs and Meth data
head(cpgLocationStatTest) #Confirm formatting
```

### All vs. Meth

```{r}
CpGLocationAllMeth <- data.frame() #Create empty dataframe to store chi-squared results
for(i in 1:ncol(cpgLocationStatTest)) { #For each genome feature
  AllFeature <- cpgLocationStatTest[1,i] #Variable for # genome feature overlaps for genomic CpGs with data
  MethFeature <- cpgLocationStatTest[2,i] #Variable for # genome feature overlaps for methylated CpGs
  AllNotFeature <- sum(cpgLocationStatTest[1,-i]) #Variable for # other CpG types for genomic CpGs with data
  MethNotFeature <- sum(cpgLocationStatTest[2,-i]) #Variable for # other CpG types for methylated CpGs
  ct <- matrix(c(AllFeature, MethFeature, AllNotFeature, MethNotFeature), ncol = 2) #Create contingency table
  colnames(ct) <- c(as.character(colnames(cpgLocationStatTest[i])), paste0("Not", colnames(cpgLocationStatTest[i]))) #Assign column names: type, not type
  rownames(ct) <- c(as.character(row.names(cpgLocationStatTest)[c(1,2)])) #Assign row names: genomic CpGs with data, methylated CpGs
  print(ct) #Confirm table is correct
  ctResults <- data.frame(broom::tidy(chisq.test(ct))) #Create dataframe storing chi-sq stats results. Use broom::tidy to extract results from test output
  ctResults$GenomeFeature <- as.character(colnames(cpgLocationStatTest)[i]) #Add CpG type to results
  CpGLocationAllMeth <- rbind(CpGLocationAllMeth, ctResults) #Add test statistics to master table
}
```

```{r}
head(CpGLocationAllMeth)
```


```{r}
CpGLocationAllMeth$p.adj <- p.adjust(CpGLocationAllMeth$p.value, method = "fdr") #Correct p-value using FDR
range(CpGLocationAllMeth$p.adj) #Look at range of p-values
head(CpGLocationAllMeth) #Confirm changes
```

```{r}
write.csv(CpGLocationAllMeth, "CpG-location-statResults.txt", quote = FALSE, row.names = TRUE) #Save statistical output
```

## Create barplot

```{r}
head(cpgFeatureOverlaps)
```

```{r}
cpgFeatureOverlapsPercents <- cpgFeatureOverlaps #Duplicate dataframe
for (i in 1:length(cpgFeatureOverlaps)) {
  cpgFeatureOverlapsPercents[,i] <- (cpgFeatureOverlapsPercents[,i] / (sum(cpgFeatureOverlapsPercents[,i]))) * 100
} #Divide every entry by sum of the column and multiply by 100 to get percentages. Do not include gene information
head(cpgFeatureOverlapsPercents) #Check calculations
```

```{r}
cpgFeatureOverlapsPercents %>% #Take dataframe
  rownames_to_column("genomeFeature") %>% #Make rownames into a genomeFeature column
  mutate(genomeFeatureOrdered = case_when(genomeFeature == "exonUTR" ~ "1exonUTR",
                                          genomeFeature == "CDS" ~ "2CDS",
                                          genomeFeature == "intron" ~ "3intron",
                                          genomeFeature == "upstream" ~ "4upstream",
                                          genomeFeature == "downstream" ~ "5downstream",
                                          genomeFeature ==  "lncRNA" ~ "6lncRNA",
                                          genomeFeature == "TE" ~ "7TE",
                                          genomeFeature == "intergenic" ~ "8intergenic")) %>% #Add numbers to the genomeFeature names so they are ordered correctly in the legend
  pivot_longer(., cols = 2:5, names_to = "category") %>% #Pivot CpG percent columns so that the dataframe is in long format
  ggplot(., aes(x = category, y = value, fill = genomeFeatureOrdered)) + #Create a plot with methylation category on the x-axis, percent CpGs on the y-axis, and color indicated by genome feature
  geom_bar(stat = "identity", color = "grey10") + #Stacked barplot with black outlines around the bars
  scale_x_discrete(name = "",
                   labels = c("All CpGs", "High", "Moderate", "Low")) + #Modiy x-axis
  scale_y_continuous(name = "% CpGs",
                     breaks = seq(0, 100, 10)) + #Modify y-axis
  scale_fill_manual(name = "Genome Feature",
                    values = plotColors,
                    labels = c("Exon UTR",
                               "CDS",
                               "Intron",
                               "Upstream Flanks",
                               "Downstream Flanks",
                               "lncRNA",
                               "TE",
                               "Intergenic")) + #Modify legend
  theme_classic(base_size = 15) #Classic theme with minimum font size specified
ggsave("figures/CpG-feature-overlaps.pdf", width = 11, height = 8.5)
```


```{r}
save.image("../../project-oyster-oa.RData") #Save RData
```

